PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-10-17 IEEEtran/paper.tex
First edit for 2nd round

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (a) generating numerically-perturbed
executions, (b) tracing floating-point computations along with
program executions, and (c) visualizing stability
evaluations. PyTracer addresses this work (a) by relying on
Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
computing libraries instrumented with Monte-Carlo arithmetic through the
Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
Python functions, modules, and classes through meta-programming, and (c) by
visualizing stability measures throughout program executions with an
interactive Plotly dashboard. \mbox{\pytracer} is available at
\mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
under MIT License.

@@

To detect numerical instabilities, \pytracer works in three main steps:
\begin{enumerate}
    \item It instruments Python programs and their module dependencies
          automatically, to save the input and output values of functions in a trace
          file.
    \item It executes the instrumented program several times using Monte-Carlo
          Arithmetic to inject perturbations in function outputs.
    \item It aggregates all trace files to compute statistics about the traced
          values, and visualize them in a GUI.
\end{enumerate}
\pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
In Step 1, \pytracer uses monkey patching to replace the functions called
by the application with a wrapper that saves their arguments and returned
values in a trace file. In Step 2, it executes the application multiple
times with Monte-Carlo Arithmetic as implemented in the
Fuzzy~\cite{kiar2020comparing} environment. It then
aggregates the resulting trace files to compute summary statistics on
traced variables. In Step 3, a Plotly dashboard provides interactive
visualizations highlighting unstable code sections.

@@

Python loads modules through an import mechanism triggered by the
\texttt{import} keyword that involves two objects: \texttt{finders} and
\texttt{loaders}. \texttt{Finders} search for packages (or
namespaces\footnote{Python distinguishes between \texttt{regular} and
    \texttt{namespace} packages. A regular package is a directory that contains an
    \_\_init\_\_.py file and potentially subdirectories (sub-packages) that contain
    themselves an \texttt{\_\_init\_\_.py} file and so on recursively. The package
    hierarchy follows those of the directory. Namespace packages introduced in
    Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not
    contain an \_\_init\_\_.py file and allow for flexible directory structures.
    Hence, parts of the package can be located in zip files, on the network, or in
    separated directories. There are no functional differences between both types of
    packages.}) on the local storage --- such as standard packages installed through
the pip package manager --- or on a distant server. \texttt{Finders} do not load
modules. Instead, they return a specification (\texttt{spec} object) that
includes information on where to find and how to load modules. \texttt{Loaders}
create module objects, initialize them, and fill import-related attributes such
as \texttt{\_\_spec\_\_}. \texttt{Loaders} then execute modules and populate
their namespace. Finally, modules are bound to their import name in the
\texttt{sys.modules} dictionary.

@@

In functional programming, iterators traverse containers lazily, meaning that
the next element in the sequence is only computed when the application uses it.
This technique allows for the manipulation of virtually infinite sequences with
finite memory. However, it implies that no complete mapping of the container
returned by an iterator exists in memory since the application computes each
element on the fly. Therefore, iterators are not serializable which implies that
\pytracer cannot save them to output traces. A workaround would be to convert
iterators into explicit containers, however, this would increase the memory
footprint significantly. \pytracer, therefore, does not instrument
iterators.

We do not expect this limitation to have a critical impact on the
applicability of \pytracer. 
In general, iterators involve simple computations that are unlikely to be
unstable. In particular, we encountered only a few iterators in our
experiments and none of them involved floating-point values. Should
instabilities occur in iterators, \pytracer would detect them through the
inputs of the functions consuming these iterators.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
In our experiments, we used MCA in Random Rounding (RR) mode, which
simulates roundoff errors by applying function $inexact$ only to the
output returned by a function. The Precision Bounding (PB) mode that applies
the $inexact$ function on inputs can
identify other types of instabilities, in particular catastrophic
cancellations, however, it is more difficult to apply to large code bases
as it often leads to execution errors that require human intervention.

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA. This can be seen as the ratio of signal over noise.
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

@@

Monte Carlo Arithmetic samples the floating-point variability of a given
program, and hence the more MCA repetitions we have, the more accurate the
distribution will be. In theory, an infinite number of MCA repetitions
should give the exact distribution. While in practice low sample sizes (3 to 30)
can give good estimates, the issue is to capture ``rare events" that could make the
computation strongly diverge from the mean. Determining the number of
MCA samples to use for a given instrumented code is still an open research question.

@@

\subsection{Interactive visualization}
The third main component of \pytracer is its interactive trace visualizer built
with Plotly Dash~\cite{plotly}, a Python framework for web visualization. Plotly
Dash abstracts the low-level details of web applications in a highly
customizable development interface allowing to offload heavy computations to a
server if necessary. Figure~\ref{fig:visu-layout} presents the general layout of
the \pytracer visualizer, divided into five sub-components:
\begin{enumerate}
    \item Timeline view: This view displays the values of each input and output
          for a given function at a given invocation. By default, \pytracer
          shows the number of significant digits computed across the traces. The
          mean and standard deviation across perturbed samples are also
          available. The visualizer uses an upper triangle for inputs and a
          lower triangle for outputs. \pytracer internally represents all values
          as NumPy arrays. In the case of non-scalar values or types that cannot
          be trivially cast into NumPy arrays, \pytracer splits the value into
          several variables. Hence, \pytracer will transform a class containing
          $n$ floating-point parameters into $n$ variables. Each variable is
          prefixed with the original value and postfixed with the name of the
          parameter. Finally, non-scalar values (vectors or matrices) are
          represented by their average and can be inspected with the Zoom view
          (see below).
    \item Gantt chart view: This view displays the call graph as a Gantt chart
          representing the instrumented functions. Each function call is
          represented with a bar. The time overlaps reflect the caller/callee
          relation. For example, in the figure, \texttt{function1} calls
          \texttt{function2} which itself calls \texttt{function4} and
          \texttt{function5}. This view provides insights into the calling context
          for each function, which may influence numerical stability.
    \item  Source code view: This view displays a local snapshot of the source
          code file of the variable hovered in the timeline view.
    \item Zoom view: This view provides details about non-scalar values. For
          example, a matrix may have instabilities around its diagonal but not
          in other regions. The visualizer also provides additional metrics such
          as array shape, infinite norm, or condition number.
    \item  Functions list view: This view allows the user to select the
          functions plotted in the timeline view.
\end{enumerate}

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on
SciPy and scikit-learn examples. Columns \textit{trace} and \textit{parse}
report on the tracing and aggregation steps with two possible outcomes:
green \valid\xspace for success and red \cross\xspace for error. Tracing errors
occur in the application as a result of 
numerical perturbations, whereas parsing errors occur when traces result from different 
control flow paths.
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples.

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the result precision for the main outputs of SciPy and scikit-learn examples. We
measured the result precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produced multiple outputs, the average value among them. Significant bits were computed on the final program output
and could therefore be computed even when parsing failed. We use the term ``result precision"
to denote the statistical variability measured in the number of significant bits, in 
contrast with the other meanings of the word precision used in the expressions
``virtual precision" or ``single/double precision'' in this paper. The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
for RR and \texttt{Bayesian Ridge Regression} for full MCA).

@@

PyTracer supports native floating-point formats as well as compound
data types, such as complex floating-point values, N-dimensional arrays,
and arbitrary Python classes. Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the visualization of the
\texttt{fft\_1D} input and output values. Results are complex values
encoded in double precision. PyTracer allows visualizing the
result precision of real (blue points) and imaginary (orange
points) parts. The top row in Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the input mean value while the bottom
row shows the standard deviation. Even though Random Rounding
(RR) was used, function inputs were perturbed since they result from
multiple intermediate computations perturbed with RR. The points with
low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal
noise introduced by MCA on the inputs is in the order of $10^{-14}$,
two orders of magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 correspond to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} shows the results of the spline evaluation with
\texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
    [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
PyTracer. We observe a significant loss of result precision in some regions, which follows
an interesting pattern that could be further investigated.
\add{This example shows that \pytracer workds on 2D inputs and the importance of
    visualizing complex data and not only reporting the values.}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/FFT/fft_x.pdf}
    \caption{Absolute value of the mean and standard deviation values for fft 1D
        inputs within RR mode (log scale). The real part is shown in blue and
        the imaginary part in orange.}
    \label{fig:fft1D_inputs}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/FFT/fft_y.pdf}
    \caption{Absolute value of the mean and standard deviation values
        for fft 1D outputs within RR mode (log scale).}
    \label{fig:fft1D_outputs}
\end{figure}

@@

Replacing \texttt{gesdd} with\texttt{gesvd} solved the SVD convergence issue.
\pytracer allows us to quickly visualize the new result precision.
Figure~\ref{fig:face_recognition_svd} shows the result precision of the eigenvalues
computed with \texttt{gesvd} in single precision. The result precision
varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. We can note that even with the \texttt{gesvd} method,
the result precision of the singular values and eigenvectors is not uniform. 
Indeed, we can see that the result precision of the lowest singular values and eigenvectors decreases 
until reaching 0 significant bits.

@@

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues such as
\href{https://github.com/google/TensorNetwork/pull/962}{github.com/google/TensorNetwork/pull/962} or
\href{https://github.com/scipy/scipy/issues/15024}{github.com/scipy/scipy/issues/15024}
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request at \href{https://github.com/scikit-learn/scikit-learn/pull/20617}{github.com/scikit-learn/scikit-learn/pull/20617}
has recently been submitted in the scikit-learn repository to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. The
unified view of all traces allows the user to quickly navigate through the
result precision of different variables.

@@

\pytracer's tracing overhead consists of (1) module instrumentation time
(initialization overhead) and (2) data write time during the execution (runtime
overhead). The initialization overhead is proportional to the number of
functions and classes traced. For example, we measured an average initialization
time of 8 seconds during our experiments with  NumPy, SciPy, and Scikit-learn.
This overhead can be reduced by excluding intensively used functions such as
\texttt{numpy.array} from the instrumentation.
Figure~\ref{fig:performance_tracing} summarizes \pytracer's tracing overhead for
SciPy and scikit-learn examples. \textit{Slowdown} represents the slowdown of
\pytracer without fuzzy measured as the ratio between the actual execution time
(without \pytracer) and the execution time using \pytracer. The \textit{Slowdown
    RR} shows the slowdown within the Fuzzy environment.
Without Fuzzy, \pytracer introduces an average slowdown of $17\times$ for
the examples running less than 8 seconds because \pytracer spends most of its
time in the initialization step. Since this step is a constant factor of the
number of functions instrumented, we also measured the instrumentation cost
itself (\textit{Slowdown (Instrumentation)}) calculated by subtracting the
initialization time to \textit{Slowdown}. We then obtain an average slowdown of
$3.5\times$. Finally, using MCA
operations (through Fuzzy) instead of the native ones increased the average
slowdown to $37\times$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/performance.pdf}
    \caption{
        \pytracer tracing overheads. \textit{Slowdown} represents the slowdown
        of \pytracer without Fuzzy measured as the ratio between the actual
        execution time (without \pytracer) and the execution time using
        \pytracer. \textit{Slowdown (Instrumentation)} show the actual
        instrumentation cost computed by subtracting the initialization step.
        The \textit{Slowdown (RR)} is the slowdown
        of \pytracer using the Fuzzy environment with the RR mode. The
        last four examples measure the initialization
        steps depending on the modules instrumented. 
        % \tristan{there's too much text in this caption, move some of it to the text, the caption should just explain the difference between the 4 slowdowns. From the text I'm not sure what is the difference between slowdown and slowdown(RR)/slowdown(MCA).}
    }
    \label{fig:performance_tracing}
\end{figure}

@@

Figure~\ref{fig:performance_parsing} shows the postprocessing time and trace
sizes produced by \pytracer. The linear correlation between the log of the size
and the log of the time shows a significant correlation (p-value $\leq 10^{-8}$
).
% \tristan{Mention linear correlation with p-value here.}

@@

\pytracer can be used with different types of numerical perturbations,
including random seeds, data noise, or stochastic arithmetic. The
use of Verificarlo to implement stochastic arithmetic through the Fuzzy
environment enables us to study the impact of numerical perturbations
introduced in non-Python dependencies while preserving the high-level view
of Python variables and functions. For instance, the SciPy environment
wraps many computation kernels written in C, such as BLAS or LAPACK, which
have a considerable impact on the stability of applications using them.

@@

Integrating \pytracer at the early stages of the development cycle would improve
the overall numerical quality of the developed code. \pytracer's automaticity
facilitates its usage and enables its integration in unit testing. For example,
\pytracer could be used in regression tests to ensure that updates do not
degrade the numerical quality of a calculation.

@@

% \gkmod{}{Note: this paragraph just reads like a list with no real
% interpretation/commentary. I think it would be better to state the strengths
% of this technique, cite examples, and then state limitations/why it isn't the
% only technique used.}
The source-to-source technique requires a rewriting of the application to modify
floating-point types and provides fine-grained control of the analyzed code.
CADNA~\cite{jezequel2008cadna} is a library for C, C++, and Fortran implementing
CESTAC~\cite{vignes1993stochastic} stochastic arithmetic.
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model
to propagate numerical errors. MCAlib~\cite{frechtling2015mcalib} is a C library
that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using
the MPFR~\cite{fousse2007mpfr} library. ADAPT \mbox{~\cite{menon2018adapt}} uses
Automatic Differentiation to compute a sensitivity profile for each variable in a
program. Finally, the work in~\cite{tang2016software} proposes a
source-to-source framework for executing targeted code in infinite and fixed
precision with and without stochastic arithmetic. The main drawback of the
source-to-source approach is its lack of scalability since rewriting large
codes can be a tedious task.

>>>>

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (a) generating numerically-perturbed
executions, (b) tracing floating-point computations along with
program executions, and (c) visualizing stability
evaluations. PyTracer addresses \add{this work} (a) by relying on
Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
computing libraries instrumented with Monte-Carlo arithmetic through the
Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
Python functions, modules, and classes through meta-programming, and (c) by
visualizing stability measures throughout program executions with an
interactive Plotly dashboard. \mbox{\pytracer} is available \add{at}
\mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
(MIT License).

@@

To detect numerical instabilities, \pytracer works in three main steps:
\begin{enumerate}
    \item Instruments Python programs and their module dependencies
          automatically, to save the input and output values of functions in a trace
          file.
    \item Executes the instrumented program several times using Monte-Carlo
          Arithmetic to inject perturbations in function outputs.
    \item Aggregates all trace files to compute statistics about the traced
          values, and visualize them in a GUI.
\end{enumerate}
\pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
In \add{S}tep 1, \pytracer uses monkey patching to replace the functions called
by the application with a wrapper that saves their arguments and returned
values in a trace file. In \add{S}tep 2, it executes the application multiple
times with Monte-Carlo Arithmetic as implemented in the
Fuzzy~\cite{kiar2020comparing} environment. It then
aggregates the resulting trace files to compute summary statistics on
traced variables. In \add{S}tep 3, a Plotly dashboard provides interactive
visualizations highlighting unstable code sections.

@@

Python loads modules through an import mechanism triggered by the
\texttt{import} keyword that involves two objects: \texttt{finders} and
\texttt{loaders}. \texttt{Finders} search for packages (or
namespaces\footnote{Python distinguishes between \texttt{regular} and
    \texttt{namespace} packages. A regular package is a directory that contains an
    \_\_init\_\_.py file and potentially subdirectories (sub-packages) that contain
    themselves an \texttt{\_\_init\_\_.py} file and so on recursively. The package
    hierarchy follows those of the directory. Namespace packages introduced in
    Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not
    contain an \_\_init\_\_.py file and allow for flexible directory structures.
    Hence, parts of the package can be located in zip files, on the network, or in
    separated directories. There are no functional differences between both types of
    packages.}) on the local storage --- such as standard packages installed through
the pip package manager --- or on a distant server. \texttt{Finders} do not load
modules. Instead, they return a specification (\texttt{spec} object) that
includes information on where to find and how to load modules. \texttt{Loaders}
create module objects, initialize them, and fill import-related attributes such
as \texttt{\_\_spec\_\_}; then execute modules and populate
their namespace. Finally, modules are bound to their import name in the
\texttt{sys.modules} dictionary.

@@

In functional programming, iterators traverse containers lazily, meaning that
the next element in the sequence is only computed when the application uses it.
This technique allows for the manipulation of virtually infinite sequences with
finite memory. However, it implies that no complete mapping of the container
returned by an iterator exists in memory since the application computes each
element on the fly. Hence, iterators are not serializable which implies that
\pytracer cannot save them to output traces. A workaround would be to convert
iterators into explicit containers but this would increase the memory
footprint significantly. \pytracer, therefore, does not instrument
iterators.

We do not expect this limitation to have a critical impact on the
applicability of \pytracer. 
Iterators generally involve simple computations that are unlikely to be
unstable. Besides, we encountered only a few iterators in our
experiments and none of them involved floating-point values. Should
instabilities occur in iterators, \pytracer would detect them through the
inputs of the functions consuming these iterators.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
In our experiments, we used MCA in Random Rounding (RR) mode, which
simulates roundoff errors by applying function $inexact$ only to the
output returned by a function. The Precision Bounding (PB) mode 
\add{that applies the $inexact$ function on inputs}
can identify other types of instabilities, in particular catastrophic
cancellations, however, it is more difficult to apply to large code bases
as it often leads to execution errors that require human intervention.

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA. This can be seen as the ratio of signal over noise.
\remove{Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
    this formula to include confidence intervals.}

@@

Monte Carlo Arithmetic samples the floating-point variability of a given
program, and hence the more MCA repetitions we have, the more accurate the
distribution will be. In theory, an infinite number of MCA repetitions
should give the exact distribution. While in practice low sample sizes (3 to 30)
can give good estimates, the issue is to capture ``rare events" 
making the computation to strongly diverge from the mean. 
Determining the number of MCA samples to use is still an open research question. 
Recent theoretical works~\cite{sohier2021confidence} on the estimation of 
significant digits allow computing an associated confidence
interval depending on the sample size and the underlying
distribution.

@@

\subsection{Interactive visualization}
The third main component of \pytracer is its interactive trace visualizer built
with Plotly Dash~\cite{plotly}, a Python framework for web visualization. Plotly
Dash abstracts the low-level details of web applications in a highly
customizable development interface allowing to offload heavy computations to a
server if necessary. Figure~\ref{fig:visu-layout} presents the general layout of
the \pytracer visualizer, divided into five sub-components:
\begin{enumerate}
    \item Timeline view: displays the values of each input and output
          for a given function at a given invocation. By default, \pytracer
          shows the number of significant digits computed across the traces. The
          mean and standard deviation across perturbed samples are also
          available. The visualizer uses an upper triangle for inputs and a
          lower triangle for outputs. \pytracer internally represents all values
          as NumPy arrays. In the case of non-scalar values or types that cannot
          be trivially cast into NumPy arrays, \pytracer splits the value into
          several variables. Hence, \pytracer will transform a class containing
          $n$ floating-point parameters into $n$ variables. Each variable is
          prefixed with the original value and postfixed with the name of the
          parameter. Finally, non-scalar values (vectors or matrices) are
          represented by their average and can be inspected with the Zoom view
          (see below).
    \item Gantt chart view: displays the call graph as a Gantt chart
          representing the instrumented functions. Each function call is
          represented with a bar. The time overlaps reflect the caller/callee
          relation. For example, in the figure, \texttt{function1} calls
          \texttt{function2} which itself calls \texttt{function4} and
          \texttt{function5}. This view provides insights into the calling context
          for each function that may influence numerical stability.
    \item  Source code view: displays a local snapshot of the source
          code file of the variable hovered in the timeline view.
    \item Zoom view: provides details about non-scalar values. For
          example, a matrix may have instabilities around its diagonal but not
          in other regions. The visualizer also provides additional metrics such
          as array shape, infinite norm, or condition number.
    \item  Functions list view: allows the user to select the
          functions plotted in the timeline view.
\end{enumerate}

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
the tracing and aggregation steps with two possible outcomes: green
\valid\xspace for success and red \cross\xspace for error. Tracing errors occur
in the application as a result of numerical perturbations, whereas parsing
errors occur when traces result from different control flow paths (CFP). 
\add{Hence, an application that consistently fails at the same instruction will
    produce traces with equivalent CFPs which can be parsed by \pytracer as the 
    Face Recognition shows.}

Overall, the
tracing and parsing of SciPy and scikit-learn examples in Random Rounding mode
was entirely successful for 32/40 examples.

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the result precision for the main outputs of SciPy and scikit-learn examples. 
We use the term ``result precision"
to denote the statistical variability measured in the number of significant bits, in 
contrast with the other meanings of the word precision used in the expressions
``virtual precision" or ``single/double precision'' in this paper. 
We measured the result precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produced multiple outputs, the average value among them. Significant bits were computed on the final program output
and could therefore be computed even when parsing failed. 
The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
for RR and \texttt{Bayesian Ridge Regression} for full MCA).

@@

PyTracer supports native floating-point formats as well as compound
data types, such as complex floating-point values, N-dimensional arrays,
and arbitrary Python classes. Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the visualization of the
\texttt{fft\_1D} input and output values \add{that} are complex values
encoded in double precision. PyTracer allows visualizing the
result precision of real (blue points) and imaginary (orange
points) parts. The top row in Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the input \add{and output} mean value while the bottom
row shows the standard deviation. Even though Random Rounding
(RR) was used, function inputs were perturbed since they result from
multiple intermediate computations perturbed with RR. The points with
low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal
noise introduced by MCA on the inputs is in the order of $10^{-14}$,
two orders of magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 correspond to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} shows the results of the spline evaluation with
\texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
    [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
PyTracer. We observe a significant loss of result precision in some regions, which follows
an interesting pattern that could be further investigated.
\add{This example shows that \pytracer works on 2D inputs and the importance of
    visualizing complex data and not only reporting the values.}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/FFT/fft_x.png}
    \caption{Absolute value of the mean and standard deviation values for fft 1D
        inputs within RR mode (log scale). The real part is shown in blue and
        the imaginary part in orange.}
    \label{fig:fft1D_inputs}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/FFT/fft_y.png}
    \caption{Absolute value of the mean and standard deviation values
        for fft 1D outputs within RR mode (log scale).}
    \label{fig:fft1D_outputs}
\end{figure}

@@

Replacing \texttt{gesdd} \add{with} \texttt{gesvd} solved the SVD convergence issue.
\pytracer allows \add{us} to quickly visualize the new result precision.
Figure~\ref{fig:face_recognition_svd} shows the result precision of the eigenvalues
computed with \texttt{gesvd} in single precision. The result precision
varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. We can note that even with the \texttt{gesvd} method,
the result precision of the singular values and eigenvectors is not uniform. 
Indeed, we can see that the result precision of the lowest singular values and eigenvectors decreases 
until reaching 0 significant bits.

@@

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues such as
\href{https://github.com/google/TensorNetwork/pull/962}{github.com/google/TensorNetwork/pull/962} or
\href{https://github.com/scipy/scipy/issues/15024}{github.com/scipy/scipy/issues/15024}
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request \add{at} \href{https://github.com/scikit-learn/scikit-learn/pull/20617}{github.com/scikit-learn/scikit-learn/pull/20617}
has recently been submitted in the scikit-learn repository to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. The
unified view of all traces allows the user to quickly navigate through the
result precision of different variables.

@@

\pytracer's tracing overhead consists of (1) module instrumentation time
(initialization overhead) and (2) data write time during the execution (runtime
overhead). The initialization overhead is proportional to the number of
functions and classes traced. For example, we measured an average initialization
time of 8 seconds during our experiments with  NumPy, SciPy, and Scikit-learn.
This overhead can be reduced by excluding intensively used functions such as
\texttt{numpy.array} from the instrumentation.
Figure~\ref{fig:performance_tracing} summarizes \pytracer's tracing overhead for
SciPy and scikit-learn examples. \textit{Instrumentation and initialization} represents the slowdown of
\pytracer without fuzzy \add{that corresponds to PyTracer used
    only for instrumenting Python programs to catch values of function inputs
    and outputs, but without instrumentation with MCA arithmetic.}
It is measured as the ratio between the actual execution time
(without \pytracer) and the execution time using \pytracer. 
The \textit{Total with RR MCA} shows the slowdown within the Fuzzy environment.
Without Fuzzy, \pytracer introduces an average slowdown of $17\times$ for
the examples running less than 8 seconds because \pytracer spends most of its
time in the initialization step. Since this step is a constant factor of the
number of functions instrumented, we also measured the instrumentation cost
itself (\textit{Instrumentation only}) calculated by subtracting the
initialization time to \textit{Instrumentation}. We then obtain an average slowdown of
$3.5\times$. Finally, using MCA
operations (through Fuzzy) instead of the native ones increased the average
slowdown to $37\times$.
Figure~\ref{fig:performance_parsing} shows the postprocessing time and trace
sizes produced by \pytracer. \add{The p-value $\leq 10^{-8}$ shows a significant
    correlation between the log of the size and the log of the time.}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/performance.png}
    \caption{
        \pytracer tracing overheads. \textit{Slowdown} represents the slowdown
        of \pytracer without Fuzzy measured as the ratio between the actual
        execution time (without \pytracer) and the execution time using
        \pytracer. \textit{Slowdown (Instrumentation)} show the actual
        instrumentation cost computed by subtracting the initialization step.
        The \textit{Slowdown (RR)} is the slowdown
        of \pytracer using the Fuzzy environment with the RR mode. The
        last four examples measure the initialization
        steps depending on the modules instrumented. 
        % \tristan{there's too much text in this caption, move some of it to the text, the caption should just explain the difference between the 4 slowdowns. From the text I'm not sure what is the difference between slowdown and slowdown(RR)/slowdown(MCA).}
    }
    \label{fig:performance_tracing}
\end{figure}

@@

% \tristan{Mention linear correlation with p-value here.}

@@

\pytracer can be used with different types of numerical perturbations,
including random seeds, data noise, or stochastic arithmetic. Using 
Verificarlo to implement stochastic arithmetic through Fuzzy
enables \add{us} to study the impact \remove{of} numerical perturbations
introduced in non-Python dependencies while preserving the high-level view
of Python variables and functions. For instance, the SciPy environment
wraps many computation kernels written in C, such as BLAS or LAPACK, which
have a considerable impact on the stability of applications using them.

@@

Integrating \pytracer at the early stages of the development cycle would improve
the overall numerical quality of the developed code. \pytracer's automaticity
facilitates its usage and enables its integration in unit testing. For example,
\pytracer could be used in regression tests to ensure that updates do not
degrade the numerical quality.

@@

The source-to-source technique requires a rewriting of the application to modify
floating-point types and provides fine-grained control of the analyzed code.
CADNA~\cite{jezequel2008cadna} is a library for C, C++, and Fortran implementing
CESTAC~\cite{vignes1993stochastic} stochastic arithmetic.
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model
to propagate numerical errors. MCAlib~\cite{frechtling2015mcalib} is a C library
that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using
the MPFR~\cite{fousse2007mpfr} library. ADAPT \mbox{~\cite{menon2018adapt}} uses
Automatic Differentiation to compute a sensitivity profile for each variable in a
program. Finally, the work in~\cite{tang2016software} propose\add{s} a
source-to-source framework for executing targeted code in infinite and fixed
precision with and without stochastic arithmetic. The main drawback of the
source-to-source approach is its lack of scalability since rewriting large
codes can be a tedious task.
